Tidy data structure to support exploration and modeling of temporal-context data

### 2018-08-20 main.Rmd
tweaking first few sections

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Mining temporal-context data for information is often inhibited by a multitude of time formats: irregular or multiple time intervals, multiple observational units or repeated measurements on multiple individuals, heterogeneous data types, nested and crossed factors indicating hierarchical sub-groups. Time series models, in particular, the software supporting time series forecasting makes strict assumptions on data that needs to be provided, typically a matrix of numeric data with an implicit time index. Going from raw data to model-ready data is painful. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are applied, and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared index variable representing time, (3) a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntax-based and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound data pipeline facilitates a succinct and transparent workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline structure and usage. The infrastructure of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal context, time series, data structure, R"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

Temporal-context data consist of observational units indexed at different time points $X_{jt}$, where the $j$\textsuperscript{th} in advance unit takes measurements of $X$ on over time $t$, for $j = 1, \dots, N_{i}$ and $1 \le t \le T$. Time primarily forms the contextual basis of temporal data, but it could arrive in many possible formats. For example, data recorded at fine time resolutions (hours, minutes, and seconds) are typically associated with different time zones and daylight savings. Temporal data often carries with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures, linking to other data tables, and etc.

@@

Time series and longitudinal data are the common terms when referring to temporal-context data. Despite of exactly the same data input, its representation diverges in these two areas due to different modelling focuses. Time series can be univariate or multivariate, but usually involves observations of relatively large in length (i.e. large $T$) possessed by each series. Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic process, for the primary purpose of forecasting, temporal dynamics, and etc. Time series are represented as vectors or matrices in statistical software, for example MATLAB [@Matlab] and R [@R-base] (including `ts`, **zoo** [@zoo2005], and **xts** [@R-xts]). The structure the multivariate time series are in is called **wide format** where each row is assumed to hold observations at a time point and each column to contain a single time series. This implies that data columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric cases. In addition to homogeneity, time indices must be stripped off the data and implicitly inferred as attributes or meta-information; the number of observations must be the same across all the series. Data wrangling from "raw" form to this specialist format can be frustrating and difficult, in order to perform a variety of downstream tasks such as analytics.

On the other hand, those who are engaged in longitudinal design, are primarily interested in explaining variations among individuals in change process and making inference about the underlying population. Longitudinal data or panel data typically refers to fewer measurements (small $T$) over a large number of individuals (large $N$). It often occurs to longitudinal design that some individuals are not observed at certain points in time, resulting in an unbalanced panel. This data issue leads representing longitudinal data much more commonly in **long format** where each row corresponds to an observational unit per time and each column contains measurements across the individuals for all points in time. Evidently, this data organisation saves storage space for many sparse cells, compared to structuring it in that wide format. Another appealing feature for a long format is data structured in a concise and semantic manner with reference to observations and variables. This supports perceiving time index as an explicit variable instead of an implicit attribute, and subsequently opens the door to manipulating time such as extracting time components. In general, fewer restrictions are made if representing temporal-context data in long format.

>>>>

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Mining temporal-context data for information is often inhibited by a multitude of time formats: irregular or multiple time intervals, multiple observational units or repeated measurements on multiple individuals, heterogeneous data types, nested and crossed factors indicating hierarchical sub-groups. Time series models, in particular, the software supporting time series forecasting makes strict assumptions on data that needs to be provided, typically a matrix of numeric data with an implicit time index. Going from raw data to model-ready data is painful. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are applied, and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared index variable representing time, (3) a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntax-based and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound data pipeline facilitates a succinct and transparent workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline structure and usage. The infrastructure of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal data, time series, data structures, data wrangling, tidy data, R, forecasting, data science, exploratory data analysis"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

Temporal-context data, $X_{jt}$, consist of $j = 1, \dots, N_{i}$ observational units indexed at different time points, $1 \le t \le T$. Time forms the contextual basis of temporal data, but it can arrive in many possible formats, which makes preparing it for modeling and forecasting can be painful. For example, data can be recorded at various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures.

@@

Despite of exactly the same data input, the representation of time series and longitudinal data diverges due to different modelling approaches.

Time series can be univariate or multivariate, and for modelling requires relatively large in length (i.e. large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterising temporal dynamics. The time series supporting modeling are represented as vectors or matrices in most statistical software, with the standards being provided by the R [@R-base] packages `ts`, **zoo** [@zoo2005], and **xts** [@R-xts] (*Why include MATLAB [@Matlab]??*). Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wide format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric variables. In addition, time indexes are stripped off the data and implicitly inferred as attributes or meta-information. It strictly requires that the number of observations must be the same across all the series. Data wrangling from the form that data arrives in, to this specialist format can be frustrating and difficult, inhibiting the variety of downstream tasks such as analytics.

For longitudinal analysis, researchers and analysts are primarily interested in explaining trends across and variations among individuals, and making inference about a broader population. Longitudinal data or panel data typically assumes fewer measurements (small $T$) over a large number of individuals (large $N$). It often occurs that measurements for individuals are taken at different time points, resulting in an unbalanced panel. Thus, the primary format required for modeling is stacked series, blocks of measurements for each individual, with columns indicating individual, time of measurement and the measurements. (The tidy data name for this would be **long format**.) Evidently, this data organisation saves storage space for many sparse cells, compared to structuring it in that wide format which would have missing values in many cells. A detriment of this format is that demographic information for subjects is often repeated for each time point. However, appealing feature is that data is structured in a concise and semantic manner with reference to observations and variables, with the time index stated explicitly. This opens the door to easily operating on time to make calculations and extract different temporal components, such as month and day of the week. It is conducive to examining the data in many different ways and leading to more comprehensive exploration and forecasting.
